Supplementary figures

### 2015-12-08 Manuscript/manuscript.tex
Switching lattice and fully-connected results
Presenting lattice results as the main results and pushing fully-connected results to the supp info, with changes to figures, captions, and the text.

The model was parameterized according to the following assumptions (Table~\ref{params}). 
(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{em}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}, though we relax this below).  
(3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=5$).  We varied infectious survival ($\nu$), probability of direct transmission ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.

The parameterization above assumes that all patches are equally accessible from all other patches (i.e. are separated by effective distance 1).  To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighbouring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).

@@

For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).

To more carefully explore the role of habitat quality in within-patch dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  
To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), infectious survival ($\nu = 0.2$), and landscape structure (fully connected).
We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.

@@

A wide range of disease dynamics were observed over the range of parameters and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by infectious survival, with high infectious survival (i.e. low disease-induced mortality) facilitating widespread infection (in which no susceptible host populations persist), and very low infectious survival (i.e. high disease-induced mortality) facilitating disease-free dynamics or intermediate spread (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with low infectious survival (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Increasing overall habitat quality facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of susceptible maintenance (Supplement).

@@

However, for pathogens falling between these extremes (e.g. with intermediate infectious survival), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap became smaller, apparent only for the most virulent pathogens.
Again, metapopulations with a lattice structure displayed similar patterns (Supplement).

To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement).  
However, because of this stability, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Supplement).
Moreover, assuming preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing the probability of extinction substantially compared to random habitat selection (0.29 vs. 0.03).

@@

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a metapopulation-level trap for a smaller range of potential pathogens (Fig~\ref{highvlow}, blue region shrinks from panels a. to c.).  
Environmentally long-lived pathogens are able to spread widely under a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
These dynamics reflect the findings of Gog \emph{et al.} (\cite{Gog2002}), who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).

@@

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_3.eps}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.  Coloured points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.  At low longevities, this pathogen is unable to invade and high quality habitat supports larger populations of susceptible hosts.  However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
\label{popsizes}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_4.eps}
\centering
\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.3$, and an envioronmental longevity equal to the expected residence time of a susceptible population on a unit quality patch (10 units of time here).  The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.  The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.  Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.  Note the overall more consistent occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
\label{simvis}
\end{figure}

>>>>

The model was parameterized according to the following assumptions (Table~\ref{params}). 
(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{em}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}, though we relax this below).  
(3)  Patches are arranged in a lattice and only accessible from their four nearest neighbors ($d_{ij}=1$ for all $i$, $j$ neighbors).  
In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=2$).  We varied infectious survival ($\nu$), probability of direct transmission ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.

To examine the effects of a less rigid spatial structure, we also implemented a model in which all patches are equally accessible from all other patches (i.e. $d_{ij} = 1$ for all $i \neq j$, in which case average migration distance was adjusted to $D = 5$ to maintain roughly equivalent occupancy).  
The lattice and fully-connected models book-end the range of possible landscape structures, with most realistic metapopulations lying somewhere in between.

@@

For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (lattice or fully connected), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).

To more carefully explore the role of habitat quality in within-patch dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  
To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.5$), infectious survival ($\nu = 0.2$), and landscape structure (lattice).
We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.

@@

A wide range of disease dynamics were observed over the range of parameters and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by infectious survival, with high infectious survival (i.e. low disease-induced mortality) facilitating widespread infection (in which no susceptible host populations persist), and very low infectious survival (i.e. high disease-induced mortality) facilitating disease-free dynamics or intermediate spread (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with low infectious survival (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Increasing overall habitat quality facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a fully-connected structure exhibited qualitatively similar patterns, but with generally easier pathogen spread (Supplement).

@@

However, for pathogens falling between these extremes (e.g. with intermediate infectious survival), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap shifted towards more virulent pathogens.
Again, metapopulations with a fully-connected structure displayed similar patterns (Supplement).

To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.5$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement).  
However, because of this stability, disease risk, measured by the mean proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Supplement).

These qualitative trends are consistent across both landscape structures, but the effects of quality and preference are more pronounced for the fully-connected case (Supplement).
In the fully connected model, high quality patches attract infectious colonists from the entire metapopulation, not just the local neighborhood, and thus are more consistently infected than in the lattice model (Supplement).

@@

Given that the presence of this ecological trap depends on the relative magnitude of pathogen spread, we find that pathogens with long-lived environmental reservoirs only generate traps when infectious survival is low enough that the pathogen is not able to spread widely.
Because environmentally long-lived pathogens are able to spread and persist more easily, increasing longevity can overwhelm the ability of low quality habitat to slow spread and provide refuges for susceptible hosts.
These dynamics reflect the findings of Gog \emph{et al.} (\cite{Gog2002}), who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).

@@

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_3.eps}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.5, respectively.  Coloured points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.  At low longevities, this pathogen is unable to invade and high quality habitat supports larger populations of susceptible hosts.  However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
\label{popsizes}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_4.eps}
\centering
\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.5$, and an envioronmental longevity of $\beta = 3.2$.  The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.  The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.  Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.  Note the overall more consistent occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
\label{simvis}
\end{figure}

### 2015-12-08 Manuscript/supplementary.tex
Switching lattice and fully-connected results
Presenting lattice results as the main results and pushing fully-connected results to the supp info, with changes to figures, captions, and the text.

Supplemental figures 1 and 2 are analogous to figures 1 and 2 from the main text, but with habitat patches arranged in a lattice structure (i.e. each habitat patch is accessible only from its four nearest neighbors).

Supplemental figures 3 and 4 show within-patch occupancy for a given $\nu$ and $\delta$.

\begin{figure}
\includegraphics[scale=0.5]{figure/supplement_1.pdf}
\caption{Median total susceptible population size as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).  Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show low quality (ranging from 0.23 to 1.24, a - c) and high quality (ranging from 0.76 to 1.77, d - f) patch quality distributions.  Darker shading corresponds to larger population sizes, while white indicates that no susceptible host populations persist (i.e. all host populations become infected).}
\label{endemic}
\end{figure}

\begin{figure}
\includegraphics[scale=0.5]{figure/supplement_2.pdf}
\centering
\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).  Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.  Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
\label{highvlow}
\end{figure}

\begin{figure}
\includegraphics[scale=0.6]{figure/supplement_3}
\centering
\caption{Probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality. Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$). Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.  Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
\end{figure}

\begin{figure}
\includegraphics[scale=0.6]{figure/supplement_4}
\centering
\caption{Probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality. Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$). Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.  Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
\end{figure}

\end{document}

>>>>

Supplemental figures 1, 2, and 3 are analogous to figures 1, 2, and 3 from the main text, but with all habitat patches equally connected (i.e. each habitat patch is equally accessible from all other habitat patches).

Supplemental figures 4 and 5, and 6 and 7 show within-patch occupancy for the lattice model and fully connected models, respectively, both with a fixed $\nu$ and $\delta$.

\begin{figure}
\includegraphics[angle = 270, scale=0.5]{figure/supplement_1.eps}
\caption{Median total susceptible population size as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).  Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show low quality (ranging from 0.23 to 1.24, a - c) and high quality (ranging from 0.76 to 1.77, d - f) patch quality distributions.  Darker shading corresponds to larger population sizes, while white indicates that no susceptible host populations persist (i.e. all host populations become infected).}
\label{endemic}
\end{figure}

\begin{figure}
\includegraphics[angle = 270, scale=0.5]{figure/supplement_2.eps}
\centering
\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).  Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.  Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
\label{highvlow}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/supplement_3.eps}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.  Coloured points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.  At low longevities, this pathogen is unable to invade and high quality habitat supports larger populations of susceptible hosts.  However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
\label{popsizes}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.6]{figure/supplement_4.eps}
\centering
\caption{Mean probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality. Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$). Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.  Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.5$, with lattice landscape structure.}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.6]{figure/supplement_5.eps}
\centering
\caption{Mean probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality. Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$). Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.  Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.5$, with lattice landscape structure.}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.6]{figure/supplement_6.eps}
\centering
\caption{Mean probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality. Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$). Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.  Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully-connected landscape structure.}
\end{figure}

\begin{figure}
\includegraphics[angle=270, scale=0.6]{figure/supplement_7.eps}
\centering
\caption{Mean probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality. Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$). Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.  Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully-connected landscape structure.}
\end{figure}
\end{document}
